---
title: "Raster Assignment"
author: "Briana Allman"
date: 10/22/2024
format: html
editor: visual
---

Load the raster and point data in and crop to an appropriate size to display the locations of the sampling plots and make a label for each site.

```{r}

library(tidyverse)
library(raster)
library(sf)
library(ggplot2)
library(ggspatial)
library(shiny)

beetle_url <- "https://raw.githubusercontent.com/DyerlabTeaching/Raster-Data/main/data/AraptusDispersalBias.csv"

bug <- read_csv( beetle_url )

raster_url <- "https://github.com/DyerlabTeaching/Raster-Data/raw/main/data/alt_22.tif"

altitude_raster <- raster(raster_url)

# Convert bug data to spatial points using sf
bug_sf <- st_as_sf(bug, coords = c("Longitude", "Latitude"), crs = 4326)

# Crop the raster to the extent of the sampling plots
cropped_raster <- crop(altitude_raster, extent(bug_sf))

# Create a base map with the raster and sampling points
ggplot() +
  layer_spatial(cropped_raster) +
  geom_sf(data = bug_sf, color = "red", size = 3) +
  geom_sf_label(data = bug_sf, aes(label = Site), nudge_y = 0.01, nudge_x = 0.01, size = 3, color = "black") +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", style = north_arrow_fancy_orienteering()) +
  ggtitle("Site Map") +
  labs(x = "Longitude", y = "Latitude") +  # Add labels for the x and y axes
  theme_minimal()+
  theme(legend.position = "none")
```

Use the `click()` function to crop the raster and filter the sites to include only the sampling locations that are in the Cape regions (e.g., south of site labeled #75).  Plot the raster of elevation and all cape region locales with sex-ratio represented as the size of the plot symbol.

```{r}
# Coordinates for site 75
site75_coords <- c(Longitude = -110.7460, Latitude = 24.58843)

# Filter for cape sites (south of site 75)
cape_sites <- bug[bug$Latitude < site75_coords["Latitude"], ]

cape_sites <- bug_sf %>% 
  filter(Latitude < site75_coords["Latitude"])

cropped_raster <- crop(altitude_raster, st_bbox(cape_sites))

```

The peninsula of Baja California is divided into the States of *Baja California Norte* and *Baja California Sur*. The border between these states is at 28° Latitude. Divide the sample locations into groups based on which state they are located in and plot the average sex ratio of the sample sites partitioned by each site.  

```{r}

```

Is there a relationship between the observed sex ratio and the elevation at that site?  Plot these, and in the text, perform a correlation test (cor.test()) and report the values as if you were writing them up in a results section.

```{r}

```
